---
title: "Pitch Prediction 2"
author: "Brian Bruxvoort"
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Required Libraries

```{r libraries}
library(tidyverse)
library(nnet)  # For multinomial logistic regression
library(caret) # For confusion matrix and model evaluation
```

## Option to Load Pre-Built Models

```{r}
pitcher_interaction_models <- readRDS("pitcher_interaction_models.rds")
pitch_data_features <- readRDS("pitch_data_features_interaction.rds")
interaction_model_evaluations <- readRDS("interaction_model_evaluations.rds")
```

```{r}
pitch_data <- read.csv("~/Truman_PDAT_630/Thesis_Project/pitch_data.csv")
```

```{r}
selected_columns2 <- c(
  "pitch_type", "player_name", "batter", "pitcher", "events", 
  "description", "zone", "stand", "p_throws", "type", "hit_location",
  "bb_type", "balls", "strikes", "pfx_x", "pfx_z", "plate_x", "plate_z",
  "outs_when_up", "inning", "sz_top", "sz_bot", "effective_speed",
  "release_spin_rate", "release_extension", "release_pos_y",
  "at_bat_number", "pitch_number", "pitch_name", "spin_axis",
  "n_thruorder_pitcher", "arm_angle")

pitch_data_clean2 <- pitch_data[, selected_columns2]
```

```{r}
selected_columns <- c(
  "pitch_type", "player_name", "zone", "stand", "balls", "strikes", "at_bat_number", "pitch_number","n_thruorder_pitcher")

pitch_data_clean <- pitch_data[, selected_columns]
```

```{r}
summary(pitch_data_clean)
```

```{r load-data}
head(pitch_data_clean)
```

## Data Preparation: Create Lagged Features and Interaction Variable

```{r create-lagged-features}
# Sort data by pitcher, at-bat, and pitch number to ensure correct sequence
pitch_data_sorted <- pitch_data_clean %>%
  arrange(player_name, at_bat_number, pitch_number)

# Create lagged features for each at-bat
pitch_data_features <- pitch_data_sorted %>%
  group_by(player_name, at_bat_number) %>%
  mutate(
    # Previous pitch features (lag 1)
    prev_pitch_type_1 = lag(pitch_type, 1),
    prev_zone_1 = lag(zone, 1),
    
    # Previous pitch features (lag 2)
    prev_pitch_type_2 = lag(pitch_type, 2),
    prev_zone_2 = lag(zone, 2),
    
    # Previous pitch features (lag 3)
    prev_pitch_type_3 = lag(pitch_type, 3),
    prev_zone_3 = lag(zone, 3)
  ) %>%
  ungroup() %>%
  # Remove first pitch of each at-bat (no previous pitch to predict from)
  filter(!is.na(prev_pitch_type_1)) %>%
  # Create interaction variable: pitch_type + zone
  mutate(
    pitch_type_zone = paste0(pitch_type, "_", zone)
  ) %>%
  # Convert categorical variables to factors
  mutate(
    pitch_type = as.factor(pitch_type),
    zone = as.factor(zone),
    pitch_type_zone = as.factor(pitch_type_zone),
    prev_pitch_type_1 = as.factor(prev_pitch_type_1),
    prev_pitch_type_2 = as.factor(prev_pitch_type_2),
    prev_pitch_type_3 = as.factor(prev_pitch_type_3),
    prev_zone_1 = as.factor(prev_zone_1),
    prev_zone_2 = as.factor(prev_zone_2),
    prev_zone_3 = as.factor(prev_zone_3),
    stand = as.factor(stand),
    balls = as.factor(balls),
    strikes = as.factor(strikes)
  )

# Check the prepared data
head(pitch_data_features)
cat("\n\nInteraction variable created: pitch_type_zone\n")
cat("Number of unique combinations:", n_distinct(pitch_data_features$pitch_type_zone), "\n")
```

## Explore Data for Model Building

```{r explore-data}
# Check number of pitches per pitcher
pitcher_counts <- pitch_data_features %>%
  group_by(player_name) %>%
  summarise(
    n_pitches = n(),
    n_pitch_types = n_distinct(pitch_type),
    n_zones = n_distinct(zone),
    n_combinations = n_distinct(pitch_type_zone)
  ) %>%
  arrange(desc(n_pitches))

print(pitcher_counts)

# Look at overall distribution of combinations
cat("\n\nOverall distribution of pitch type + zone combinations:\n")
combination_counts <- pitch_data_features %>%
  count(pitch_type_zone, sort = TRUE) %>%
  head(20)
print(combination_counts)

# Filter pitchers with sufficient data (at least 200 pitches for interaction model)
# Need significantly more data because we're predicting many more outcome classes
pitchers_sufficient_data <- pitcher_counts %>%
  filter(n_pitches >= 200) %>%
  pull(player_name)

cat("\nNumber of pitchers with sufficient data (>=200 pitches):", length(pitchers_sufficient_data))
```

## Build Multinomial Logistic Regression Models (Interaction Model)

```{r build-models}
# # Function to build INTERACTION model (pitch_type + zone combined)
# build_pitcher_interaction_model <- function(pitcher_data) {
#   tryCatch({
#     # Check if there are enough distinct combinations
#     n_combinations <- n_distinct(pitcher_data$pitch_type_zone)
#     n_pitches <- nrow(pitcher_data)
#     
#     cat(" | Combinations:", n_combinations, "| Pitches:", n_pitches)
#     
#    # Need at least 5 observations per combination on average
#     if (n_pitches / n_combinations < 5) {
#       cat(" - Too sparse")
#       return(NULL)
#     }
#     
#     # Filter out very rare combinations (less than 3 occurrences)
#     combination_counts <- pitcher_data %>%
#       count(pitch_type_zone) %>%
#       filter(n >= 3)
#     
#     # Keep only common combinations
#     pitcher_data_filtered <- pitcher_data %>%
#       filter(pitch_type_zone %in% combination_counts$pitch_type_zone) %>%
#       mutate(pitch_type_zone = droplevels(pitch_type_zone))
#     
#     cat(" | Filtered to:", n_distinct(pitcher_data_filtered$pitch_type_zone), "combinations")
#     
#     # Check if still have enough data
#     if (nrow(pitcher_data_filtered) < 50) {
#       cat(" - Insufficient data after filtering")
#       return(NULL)
#     }
#     
#     # Build multinomial logistic regression model for pitch_type_zone interaction
#     # Use simpler model with just lag 1 to reduce complexity
#     model <- multinom(
#       pitch_type_zone ~ prev_pitch_type_1 + prev_zone_1 + 
#                         balls + strikes + stand,
#       data = pitcher_data_filtered,
#       maxit = 1000,
#       trace = FALSE,
#       MaxNWts = 5000  # Increase weight limit for complex models
#     )
#     return(model)
#   }, error = function(e) {
#     cat(" - Error:", e$message)
#     return(NULL)
#   })
# }
# 
# # Build interaction models for each pitcher
# pitcher_interaction_models <- list()
# 
# for (pitcher in pitchers_sufficient_data) {
#   cat("\nBuilding interaction model for:", pitcher)
#   
#   pitcher_data <- pitch_data_features %>%
#     filter(player_name == pitcher)
#   
#   model <- build_pitcher_interaction_model(pitcher_data)
#   
#   if (!is.null(model)) {
#     pitcher_interaction_models[[pitcher]] <- model
#     cat(" - Success")
#   } else {
#     cat(" - Failed")
#   }
# }
# 
# cat("\n\nTotal interaction models built:", length(pitcher_interaction_models))
```

## Evaluate Model Performance

```{r evaluate-models}
# Function to evaluate a pitcher's interaction model
evaluate_pitcher_interaction_model <- function(pitcher_name, model, pitcher_data) {
  tryCatch({
    # Make predictions
    predictions <- predict(model, newdata = pitcher_data, type = "class")
    
    # Calculate accuracy
    accuracy <- mean(as.character(predictions) == as.character(pitcher_data$pitch_type_zone), na.rm = TRUE)
    
    # Also calculate type-only and zone-only accuracy from interaction predictions
    # Extract type and zone from predictions
    pred_split <- str_split_fixed(as.character(predictions), "_", 2)
    actual_split <- str_split_fixed(as.character(pitcher_data$pitch_type_zone), "_", 2)
    
    type_accuracy <- mean(pred_split[,1] == actual_split[,1], na.rm = TRUE)
    zone_accuracy <- mean(pred_split[,2] == actual_split[,2], na.rm = TRUE)
    
    # Create confusion matrix
    conf_matrix <- tryCatch({
      pred_char <- as.character(predictions)
      actual_char <- as.character(pitcher_data$pitch_type_zone)
      common_levels <- sort(unique(c(pred_char, actual_char)))
      
      confusionMatrix(
        factor(pred_char, levels = common_levels),
        factor(actual_char, levels = common_levels)
      )
    }, error = function(e) NULL)
    
    return(list(
      pitcher = pitcher_name,
      combined_accuracy = accuracy,
      type_accuracy = type_accuracy,
      zone_accuracy = zone_accuracy,
      confusion_matrix = conf_matrix,
      error = NULL
    ))
  }, error = function(e) {
    return(list(
      pitcher = pitcher_name,
      combined_accuracy = NA,
      type_accuracy = NA,
      zone_accuracy = NA,
      confusion_matrix = NULL,
      error = e$message
    ))
  })
}

# Evaluate all interaction models
interaction_model_evaluations <- list()

for (pitcher in names(pitcher_interaction_models)) {
  pitcher_data <- pitch_data_features %>%
    filter(player_name == pitcher)
  
  eval <- evaluate_pitcher_interaction_model(pitcher, pitcher_interaction_models[[pitcher]], pitcher_data)
  interaction_model_evaluations[[pitcher]] <- eval
}

# Summary of interaction model accuracies
interaction_accuracy_summary <- data.frame(
  pitcher = names(interaction_model_evaluations),
  combined_accuracy = sapply(interaction_model_evaluations, function(x) x$combined_accuracy),
  type_accuracy = sapply(interaction_model_evaluations, function(x) x$type_accuracy),
  zone_accuracy = sapply(interaction_model_evaluations, function(x) x$zone_accuracy),
  error = sapply(interaction_model_evaluations, function(x) {
    if (!is.null(x$error)) return(x$error) else return(NA)
  })
) %>%
  arrange(desc(combined_accuracy))

print(interaction_accuracy_summary)

# Summary statistics
cat("\n\n=== INTERACTION MODEL PERFORMANCE ===\n")
cat("Models with errors:", sum(!is.na(interaction_accuracy_summary$error)), "\n")
cat("Models successful:", sum(is.na(interaction_accuracy_summary$error) & !is.na(interaction_accuracy_summary$combined_accuracy)), "\n")
cat("Mean combined accuracy:", round(mean(interaction_accuracy_summary$combined_accuracy, na.rm = TRUE), 4), "\n")
cat("Mean type accuracy (from interaction):", round(mean(interaction_accuracy_summary$type_accuracy, na.rm = TRUE), 4), "\n")
cat("Mean zone accuracy (from interaction):", round(mean(interaction_accuracy_summary$zone_accuracy, na.rm = TRUE), 4), "\n")
```

```{r}
# # Save the models and data for future use
# cat("\n\nSaving models and data...\n")
# saveRDS(pitcher_interaction_models, file = "pitcher_interaction_models.rds")
# saveRDS(pitch_data_features, file = "pitch_data_features_interaction.rds")
# saveRDS(interaction_model_evaluations, file = "interaction_model_evaluations.rds")
# cat("Models saved successfully!\n")
# cat("Files created:\n")
# cat("  - pitcher_interaction_models.rds\n")
# cat("  - pitch_data_features_interaction.rds\n")
# cat("  - interaction_model_evaluations.rds\n")
```

## Example: Detailed Results for One Pitcher

```{r example-pitcher}
# Select the first pitcher with a model
example_pitcher <- names(pitcher_interaction_models)[1]

cat("Detailed results for:", example_pitcher, "\n\n")

# Show basic model info (much faster than full summary)
cat("Model Information:\n")
model <- pitcher_interaction_models[[example_pitcher]]
cat("Number of outcome classes:", length(model$lev), "\n")
cat("Outcome classes (first 10):", paste(head(model$lev, 10), collapse = ", "), "\n")
cat("Number of predictors:", length(model$coefnames), "\n")
cat("Convergence:", ifelse(model$convergence == 0, "Yes", "No"), "\n")

# Show confusion matrix
cat("\n\nConfusion Matrix:\n")
print(interaction_model_evaluations[[example_pitcher]]$confusion_matrix)

# Get just the first 3 rows for example predictions
example_data <- pitch_data_features %>%
  filter(player_name == example_pitcher) %>%
  head(3)  # Only get first 3 pitches

# Make predictions for just these 3 pitches
predicted_probs <- predict(pitcher_interaction_models[[example_pitcher]], 
                           newdata = example_data, 
                           type = "probs")

# Show predictions with probabilities
cat("\n\nExample Predictions (first 3 pitches):\n")
cat("Top 5 most likely pitch+zone combinations for each:\n\n")

for (i in 1:nrow(example_data)) {
  cat("Pitch", i, ":\n")
  cat("Actual:", as.character(example_data$pitch_type_zone[i]), "\n")
  
  # Get top 5 predictions
  probs_i <- predicted_probs[i, ]
  top_5 <- sort(probs_i, decreasing = TRUE)[1:5]
  
  for (j in 1:5) {
    cat(sprintf("  %d. %s: %.1f%%\n", j, names(top_5)[j], top_5[j] * 100))
  }
  cat("\n")
}
```

## Predict Next Pitch Type AND Zone (Interaction Model)

```{r predict-function}
# Function to predict next pitch using interaction model
predict_next_pitch_interaction <- function(pitcher_name, prev_pitch_1, prev_zone_1, 
                                           prev_pitch_2 = NA, prev_zone_2 = NA,
                                           balls, strikes, stand) {
  
  if (!pitcher_name %in% names(pitcher_interaction_models)) {
    return("No model available for this pitcher")
  }
  
  # Create new data frame for prediction
  # Note: Only using lag 1 since that's what the model uses
  new_data <- data.frame(
    prev_pitch_type_1 = factor(prev_pitch_1, levels = levels(pitch_data_features$prev_pitch_type_1)),
    prev_zone_1 = factor(prev_zone_1, levels = levels(pitch_data_features$prev_zone_1)),
    balls = factor(balls, levels = levels(pitch_data_features$balls)),
    strikes = factor(strikes, levels = levels(pitch_data_features$strikes)),
    stand = factor(stand, levels = levels(pitch_data_features$stand))
  )
  
  # Get predictions
  tryCatch({
    predicted_combination <- predict(pitcher_interaction_models[[pitcher_name]], newdata = new_data, type = "class")
    combination_probs <- predict(pitcher_interaction_models[[pitcher_name]], newdata = new_data, type = "probs")
    
    # Parse the combination
    combination_split <- str_split_fixed(as.character(predicted_combination), "_", 2)
    predicted_type <- combination_split[1]
    predicted_zone <- combination_split[2]
    
    # Get top 10 most likely combinations
    top_probs <- sort(combination_probs, decreasing = TRUE)[1:min(10, length(combination_probs))]
    
    return(list(
      predicted_pitch_type = predicted_type,
      predicted_zone = predicted_zone,
      predicted_combination = as.character(predicted_combination),
      top_probabilities = top_probs
    ))
  }, error = function(e) {
    return(paste("Prediction error:", e$message))
  })
}

# Example prediction
example_prediction <- predict_next_pitch_interaction(
  pitcher_name = example_pitcher,
  prev_pitch_1 = "FF",  # Previous pitch was a fastball
  prev_zone_1 = "2",    # In zone 2
  prev_pitch_2 = "CH",  # Two pitches ago was a changeup
  prev_zone_2 = "8",    # In zone 8
  balls = 1,
  strikes = 1,
  stand = "R"           # Right-handed batter
)

cat("\n\nExample Prediction:\n")
cat("Predicted pitch type:", example_prediction$predicted_pitch_type, "\n")
cat("Predicted zone:", example_prediction$predicted_zone, "\n")
cat("Full combination:", example_prediction$predicted_combination, "\n\n")
cat("Top 10 most likely pitch+zone combinations:\n")
for (i in 1:length(example_prediction$top_probabilities)) {
  cat(sprintf("%2d. %s: %.1f%%\n", i, 
              names(example_prediction$top_probabilities)[i], 
              example_prediction$top_probabilities[i] * 100))
}
```

## Visualize Results

```{r visualizations}
# Plot COMBINED accuracy distribution
ggplot(interaction_accuracy_summary, aes(x = reorder(pitcher, combined_accuracy), y = combined_accuracy)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Interaction Model Combined Accuracy by Pitcher",
       subtitle = "Predicting exact pitch type + zone combination",
       x = "Pitcher",
       y = "Combined Accuracy") +
  theme_minimal() +
  geom_hline(yintercept = mean(interaction_accuracy_summary$combined_accuracy, na.rm = TRUE), 
             linetype = "dashed", color = "red") +
  annotate("text", x = 1, y = mean(interaction_accuracy_summary$combined_accuracy, na.rm = TRUE) + 0.02,
           label = paste0("Mean: ", round(mean(interaction_accuracy_summary$combined_accuracy, na.rm = TRUE), 3)),
           color = "red")

# Compare type vs zone accuracy from interaction model
ggplot(interaction_accuracy_summary, aes(x = type_accuracy, y = zone_accuracy)) +
  geom_point(size = 3, alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Type Accuracy vs. Zone Accuracy (from Interaction Model)",
       x = "Type Accuracy",
       y = "Zone Accuracy") +
  theme_minimal()

# Plot all three metrics together
interaction_accuracy_long <- interaction_accuracy_summary %>%
  select(pitcher, combined_accuracy, type_accuracy, zone_accuracy) %>%
  pivot_longer(cols = c(combined_accuracy, type_accuracy, zone_accuracy),
               names_to = "metric",
               values_to = "accuracy") %>%
  mutate(metric = factor(metric, 
                        levels = c("combined_accuracy", "type_accuracy", "zone_accuracy"),
                        labels = c("Combined (Type+Zone)", "Type Only", "Zone Only")))

ggplot(interaction_accuracy_long, aes(x = accuracy, fill = metric)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  facet_wrap(~ metric, ncol = 1) +
  labs(title = "Distribution of Accuracies from Interaction Model",
       x = "Accuracy",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("purple", "steelblue", "darkgreen"))

# Compare to baseline
baseline_accuracy <- pitch_data_features %>%
  group_by(player_name) %>%
  summarise(
    most_common_combination = names(which.max(table(pitch_type_zone))),
    baseline_combined_accuracy = max(table(pitch_type_zone)) / n()
  )

comparison <- interaction_accuracy_summary %>%
  left_join(baseline_accuracy, by = c("pitcher" = "player_name")) %>%
  mutate(improvement = combined_accuracy - baseline_combined_accuracy)

ggplot(comparison, aes(x = baseline_combined_accuracy, y = combined_accuracy)) +
  geom_point(size = 3, alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Interaction Model Combined Accuracy vs. Baseline",
       x = "Baseline Accuracy (Most Common Combination)",
       y = "Model Combined Accuracy") +
  theme_minimal()
```

## Session Info

```{r session-info}
sessionInfo()
```
