---
title: "Pitch Prediction 2"
author: "Brian Bruxvoort"
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load Required Libraries

```{r libraries}
library(tidyverse)
library(nnet)  # For multinomial logistic regression
library(caret) # For confusion matrix and model evaluation
```

## Option to Load Pre-Built Models

```{r}
pitch_data <- read.csv("~/Truman_PDAT_630/Thesis_Project/pitch_data.csv")
```

```{r}
selected_columns2 <- c(
  "pitch_type", "player_name", "batter", "pitcher", "events", 
  "description", "zone", "stand", "p_throws", "type", "hit_location",
  "bb_type", "balls", "strikes", "pfx_x", "pfx_z", "plate_x", "plate_z",
  "outs_when_up", "inning", "sz_top", "sz_bot", "effective_speed",
  "release_spin_rate", "release_extension", "release_pos_y",
  "at_bat_number", "pitch_number", "pitch_name", "spin_axis",
  "n_thruorder_pitcher", "arm_angle")

pitch_data_clean2 <- pitch_data[, selected_columns2]
```

```{r}
selected_columns <- c(
  "pitch_type", "player_name", "zone", "stand", "balls", "strikes", "at_bat_number", "pitch_number","n_thruorder_pitcher")

pitch_data_clean <- pitch_data[, selected_columns]
```

```{r}
summary(pitch_data_clean)
```

```{r load-data}
head(pitch_data_clean)
```

## Data Preparation: Create Lagged Features

```{r create-lagged-features}
# Sort data by pitcher, at-bat, and pitch number to ensure correct sequence
pitch_data_sorted <- pitch_data_clean %>%
  arrange(player_name, at_bat_number, pitch_number)

# Create lagged features for each at-bat
pitch_data_features <- pitch_data_sorted %>%
  group_by(player_name, at_bat_number) %>%
  mutate(
    # Previous pitch features (lag 1)
    prev_pitch_type_1 = lag(pitch_type, 1),
    prev_zone_1 = lag(zone, 1),
    
    # Previous pitch features (lag 2)
    prev_pitch_type_2 = lag(pitch_type, 2),
    prev_zone_2 = lag(zone, 2),
    
    # Previous pitch features (lag 3)
    prev_pitch_type_3 = lag(pitch_type, 3),
    prev_zone_3 = lag(zone, 3)
  ) %>%
  ungroup() %>%
  # Remove first pitch of each at-bat (no previous pitch to predict from)
  filter(!is.na(prev_pitch_type_1)) %>%
  # Convert categorical variables to factors
  mutate(
    pitch_type = as.factor(pitch_type),
    prev_pitch_type_1 = as.factor(prev_pitch_type_1),
    prev_pitch_type_2 = as.factor(prev_pitch_type_2),
    prev_pitch_type_3 = as.factor(prev_pitch_type_3),
    prev_zone_1 = as.factor(prev_zone_1),
    prev_zone_2 = as.factor(prev_zone_2),
    prev_zone_3 = as.factor(prev_zone_3),
    zone = as.factor(zone),
    stand = as.factor(stand),
    balls = as.factor(balls),
    strikes = as.factor(strikes)
  )

# Check the prepared data
head(pitch_data_features)
```

## Explore Data for Model Building

```{r explore-data}
# Check number of pitches per pitcher
pitcher_counts <- pitch_data_features %>%
  group_by(player_name) %>%
  summarise(
    n_pitches = n(),
    n_pitch_types = n_distinct(pitch_type)
  ) %>%
  arrange(desc(n_pitches))

print(pitcher_counts)

# Filter pitchers with sufficient data (at least 50 pitches)
pitchers_sufficient_data <- pitcher_counts %>%
  filter(n_pitches >= 50) %>%
  pull(player_name)

cat("\nNumber of pitchers with sufficient data:", length(pitchers_sufficient_data))
```

## Build Multinomial Logistic Regression Models (Type AND Location)

```{r build-models}
# # Function to build PITCH TYPE model for a single pitcher
# build_pitcher_type_model <- function(pitcher_data) {
#   tryCatch({
#     # Build multinomial logistic regression model for PITCH TYPE
#     model <- multinom(
#       pitch_type ~ prev_pitch_type_1 + prev_zone_1 +
#                    prev_pitch_type_2 + prev_zone_2 +
#                    balls + strikes + stand,
#       data = pitcher_data,
#       maxit = 500,
#       trace = FALSE
#     )
#     return(model)
#   }, error = function(e) {
#     return(NULL)
#   })
# }
# 
# # Function to build ZONE (location) model for a single pitcher
# build_pitcher_zone_model <- function(pitcher_data) {
#   tryCatch({
#     # Build multinomial logistic regression model for ZONE (location)
#     model <- multinom(
#       zone ~ prev_pitch_type_1 + prev_zone_1 +
#              prev_pitch_type_2 + prev_zone_2 +
#              balls + strikes + stand,
#       data = pitcher_data,
#       maxit = 500,
#       trace = FALSE
#     )
#     return(model)
#   }, error = function(e) {
#     return(NULL)
#   })
# }
# 
# # Build both models for each pitcher
# pitcher_type_models <- list()
# pitcher_zone_models <- list()
# 
# for (pitcher in pitchers_sufficient_data) {
#   cat("\nBuilding models for:", pitcher)
# 
#   pitcher_data <- pitch_data_features %>%
#     filter(player_name == pitcher)
# 
#   # Build pitch type model
#   type_model <- build_pitcher_type_model(pitcher_data)
#   if (!is.null(type_model)) {
#     pitcher_type_models[[pitcher]] <- type_model
#     cat(" - Type: Success")
#   } else {
#     cat(" - Type: Failed")
#   }
# 
#   # Build zone model
#   zone_model <- build_pitcher_zone_model(pitcher_data)
#   if (!is.null(zone_model)) {
#     pitcher_zone_models[[pitcher]] <- zone_model
#     cat(" | Zone: Success")
#   } else {
#     cat(" | Zone: Failed")
#   }
# }
# 
# cat("\n\nTotal pitch type models built:", length(pitcher_type_models))
# cat("\nTotal zone models built:", length(pitcher_zone_models))
```

## Evaluate Model Performance

```{r evaluate-models}
# Function to evaluate a pitcher's PITCH TYPE model
evaluate_pitcher_type_model <- function(pitcher_name, model, pitcher_data) {
  tryCatch({
    predictions <- predict(model, newdata = pitcher_data, type = "class")
    accuracy <- mean(as.character(predictions) == as.character(pitcher_data$pitch_type), na.rm = TRUE)
    
    conf_matrix <- tryCatch({
      pred_char <- as.character(predictions)
      actual_char <- as.character(pitcher_data$pitch_type)
      common_levels <- sort(unique(c(pred_char, actual_char)))
      
      confusionMatrix(
        factor(pred_char, levels = common_levels),
        factor(actual_char, levels = common_levels)
      )
    }, error = function(e) NULL)
    
    return(list(
      pitcher = pitcher_name,
      accuracy = accuracy,
      confusion_matrix = conf_matrix,
      error = NULL
    ))
  }, error = function(e) {
    return(list(
      pitcher = pitcher_name,
      accuracy = NA,
      confusion_matrix = NULL,
      error = e$message
    ))
  })
}

# Function to evaluate a pitcher's ZONE model
evaluate_pitcher_zone_model <- function(pitcher_name, model, pitcher_data) {
  tryCatch({
    predictions <- predict(model, newdata = pitcher_data, type = "class")
    accuracy <- mean(as.character(predictions) == as.character(pitcher_data$zone), na.rm = TRUE)
    
    conf_matrix <- tryCatch({
      pred_char <- as.character(predictions)
      actual_char <- as.character(pitcher_data$zone)
      common_levels <- sort(unique(c(pred_char, actual_char)))
      
      confusionMatrix(
        factor(pred_char, levels = common_levels),
        factor(actual_char, levels = common_levels)
      )
    }, error = function(e) NULL)
    
    return(list(
      pitcher = pitcher_name,
      accuracy = accuracy,
      confusion_matrix = conf_matrix,
      error = NULL
    ))
  }, error = function(e) {
    return(list(
      pitcher = pitcher_name,
      accuracy = NA,
      confusion_matrix = NULL,
      error = e$message
    ))
  })
}

# Evaluate all type models
type_model_evaluations <- list()
for (pitcher in names(pitcher_type_models)) {
  pitcher_data <- pitch_data_features %>%
    filter(player_name == pitcher)
  
  eval <- evaluate_pitcher_type_model(pitcher, pitcher_type_models[[pitcher]], pitcher_data)
  type_model_evaluations[[pitcher]] <- eval
}

# Evaluate all zone models
zone_model_evaluations <- list()
for (pitcher in names(pitcher_zone_models)) {
  pitcher_data <- pitch_data_features %>%
    filter(player_name == pitcher)
  
  eval <- evaluate_pitcher_zone_model(pitcher, pitcher_zone_models[[pitcher]], pitcher_data)
  zone_model_evaluations[[pitcher]] <- eval
}

# Summary of PITCH TYPE model accuracies
type_accuracy_summary <- data.frame(
  pitcher = names(type_model_evaluations),
  type_accuracy = sapply(type_model_evaluations, function(x) x$accuracy),
  type_error = sapply(type_model_evaluations, function(x) {
    if (!is.null(x$error)) return(x$error) else return(NA)
  })
)

# Summary of ZONE model accuracies
zone_accuracy_summary <- data.frame(
  pitcher = names(zone_model_evaluations),
  zone_accuracy = sapply(zone_model_evaluations, function(x) x$accuracy),
  zone_error = sapply(zone_model_evaluations, function(x) {
    if (!is.null(x$error)) return(x$error) else return(NA)
  })
)

# Combined summary
combined_accuracy_summary <- type_accuracy_summary %>%
  full_join(zone_accuracy_summary, by = "pitcher") %>%
  arrange(desc(type_accuracy))

print(combined_accuracy_summary)

# Summary statistics
cat("\n\n=== PITCH TYPE MODEL PERFORMANCE ===\n")
cat("Models with errors:", sum(!is.na(type_accuracy_summary$type_error)), "\n")
cat("Models successful:", sum(is.na(type_accuracy_summary$type_error) & !is.na(type_accuracy_summary$type_accuracy)), "\n")
cat("Mean accuracy:", round(mean(type_accuracy_summary$type_accuracy, na.rm = TRUE), 4), "\n")

cat("\n=== ZONE MODEL PERFORMANCE ===\n")
cat("Models with errors:", sum(!is.na(zone_accuracy_summary$zone_error)), "\n")
cat("Models successful:", sum(is.na(zone_accuracy_summary$zone_error) & !is.na(zone_accuracy_summary$zone_accuracy)), "\n")
cat("Mean accuracy:", round(mean(zone_accuracy_summary$zone_accuracy, na.rm = TRUE), 4), "\n")
```

## Save Models for Future Use
```{r save-separate-models}
# # Save the separate type and zone models
# cat("Saving type and zone models...\n")
# saveRDS(pitcher_type_models, file = "pitcher_type_models_separate.rds")
# saveRDS(pitcher_zone_models, file = "pitcher_zone_models_separate.rds")
# saveRDS(pitch_data_features, file = "pitch_data_features_separate.rds")
# saveRDS(type_model_evaluations, file = "type_model_evaluations_separate.rds")
# saveRDS(zone_model_evaluations, file = "zone_model_evaluations_separate.rds")
# 
# cat("\nModels saved successfully!\n")
# cat("Files created:\n")
# cat("  - pitcher_type_models_separate.rds\n")
# cat("  - pitcher_zone_models_separate.rds\n")
# cat("  - pitch_data_features_separate.rds\n")
# cat("  - type_model_evaluations_separate.rds\n")
# cat("  - zone_model_evaluations_separate.rds\n")
```

## Example: Detailed Results for One Pitcher

```{r example-pitcher}
# Select the first pitcher with both models
example_pitcher <- intersect(names(pitcher_type_models), names(pitcher_zone_models))[1]

cat("Detailed results for:", example_pitcher, "\n\n")

# PITCH TYPE MODEL
cat("========== PITCH TYPE MODEL ==========\n")
cat("Model Coefficients:\n")
summary(pitcher_type_models[[example_pitcher]])

cat("\n\nConfusion Matrix:\n")
print(type_model_evaluations[[example_pitcher]]$confusion_matrix)

# ZONE MODEL
cat("\n\n========== ZONE MODEL ==========\n")
cat("Model Coefficients:\n")
summary(pitcher_zone_models[[example_pitcher]])

cat("\n\nConfusion Matrix:\n")
print(zone_model_evaluations[[example_pitcher]]$confusion_matrix)

# Get predictions with probabilities
example_data <- pitch_data_features %>%
  filter(player_name == example_pitcher)

type_predicted_probs <- predict(pitcher_type_models[[example_pitcher]], 
                                newdata = example_data, 
                                type = "probs")

zone_predicted_probs <- predict(pitcher_zone_models[[example_pitcher]], 
                                newdata = example_data, 
                                type = "probs")

# Show first few predictions with probabilities
cat("\n\nExample Predictions (first 5 pitches):\n")
cat("\nPITCH TYPE probabilities:\n")
print(type_predicted_probs[1:5, ])
cat("\nZONE probabilities:\n")
print(zone_predicted_probs[1:5, ])
```

## Predict Next Pitch Type AND Zone for New Sequence

```{r predict-function}
# Function to predict next pitch TYPE AND ZONE given previous sequence
predict_next_pitch <- function(pitcher_name, prev_pitch_1, prev_zone_1, 
                                prev_pitch_2 = NA, prev_zone_2 = NA,
                                balls, strikes, stand) {
  
  # Check if models exist
  has_type_model <- pitcher_name %in% names(pitcher_type_models)
  has_zone_model <- pitcher_name %in% names(pitcher_zone_models)
  
  if (!has_type_model && !has_zone_model) {
    return("No models available for this pitcher")
  }
  
  # Create new data frame for prediction
  new_data <- data.frame(
    prev_pitch_type_1 = factor(prev_pitch_1, levels = levels(pitch_data_features$prev_pitch_type_1)),
    prev_zone_1 = factor(prev_zone_1, levels = levels(pitch_data_features$prev_zone_1)),
    prev_pitch_type_2 = factor(prev_pitch_2, levels = levels(pitch_data_features$prev_pitch_type_2)),
    prev_zone_2 = factor(prev_zone_2, levels = levels(pitch_data_features$prev_zone_2)),
    balls = factor(balls, levels = levels(pitch_data_features$balls)),
    strikes = factor(strikes, levels = levels(pitch_data_features$strikes)),
    stand = factor(stand, levels = levels(pitch_data_features$stand))
  )
  
  result <- list()
  
  # Predict pitch type
  if (has_type_model) {
    tryCatch({
      predicted_type <- predict(pitcher_type_models[[pitcher_name]], newdata = new_data, type = "class")
      type_probs <- predict(pitcher_type_models[[pitcher_name]], newdata = new_data, type = "probs")
      
      result$predicted_pitch_type <- as.character(predicted_type)
      result$type_probabilities <- type_probs
    }, error = function(e) {
      result$predicted_pitch_type <- paste("Error:", e$message)
      result$type_probabilities <- NULL
    })
  } else {
    result$predicted_pitch_type <- "No type model available"
    result$type_probabilities <- NULL
  }
  
  # Predict zone
  if (has_zone_model) {
    tryCatch({
      predicted_zone <- predict(pitcher_zone_models[[pitcher_name]], newdata = new_data, type = "class")
      zone_probs <- predict(pitcher_zone_models[[pitcher_name]], newdata = new_data, type = "probs")
      
      result$predicted_zone <- as.character(predicted_zone)
      result$zone_probabilities <- zone_probs
    }, error = function(e) {
      result$predicted_zone <- paste("Error:", e$message)
      result$zone_probabilities <- NULL
    })
  } else {
    result$predicted_zone <- "No zone model available"
    result$zone_probabilities <- NULL
  }
  
  return(result)
}

# Example prediction
example_prediction <- predict_next_pitch(
  pitcher_name = example_pitcher,
  prev_pitch_1 = "FF",  # Previous pitch was a fastball
  prev_zone_1 = "2",    # In zone 2
  prev_pitch_2 = "CH",  # Two pitches ago was a changeup
  prev_zone_2 = "8",    # In zone 8
  balls = 1,
  strikes = 1,
  stand = "R"           # Right-handed batter
)

cat("\n\nExample Prediction:\n")
cat("Predicted pitch type:", example_prediction$predicted_pitch_type, "\n")
cat("Predicted zone:", example_prediction$predicted_zone, "\n\n")
cat("Type probabilities:\n")
print(example_prediction$type_probabilities)
cat("\nZone probabilities:\n")
print(example_prediction$zone_probabilities)
```

## Visualize Results

```{r visualizations}
# Plot PITCH TYPE accuracy distribution
ggplot(combined_accuracy_summary, aes(x = reorder(pitcher, type_accuracy), y = type_accuracy)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Pitch Type Model Accuracy by Pitcher",
       x = "Pitcher",
       y = "Accuracy") +
  theme_minimal() +
  geom_hline(yintercept = mean(combined_accuracy_summary$type_accuracy, na.rm = TRUE), 
             linetype = "dashed", color = "red") +
  annotate("text", x = 1, y = mean(combined_accuracy_summary$type_accuracy, na.rm = TRUE) + 0.02,
           label = paste0("Mean: ", round(mean(combined_accuracy_summary$type_accuracy, na.rm = TRUE), 3)),
           color = "red")

# Plot ZONE accuracy distribution
ggplot(combined_accuracy_summary, aes(x = reorder(pitcher, zone_accuracy), y = zone_accuracy)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Zone Model Accuracy by Pitcher",
       x = "Pitcher",
       y = "Accuracy") +
  theme_minimal() +
  geom_hline(yintercept = mean(combined_accuracy_summary$zone_accuracy, na.rm = TRUE), 
             linetype = "dashed", color = "red") +
  annotate("text", x = 1, y = mean(combined_accuracy_summary$zone_accuracy, na.rm = TRUE) + 0.02,
           label = paste0("Mean: ", round(mean(combined_accuracy_summary$zone_accuracy, na.rm = TRUE), 3)),
           color = "red")

# Compare type vs zone accuracy
ggplot(combined_accuracy_summary, aes(x = type_accuracy, y = zone_accuracy)) +
  geom_point(size = 3, alpha = 0.6, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Pitch Type Accuracy vs. Zone Accuracy",
       x = "Pitch Type Accuracy",
       y = "Zone Accuracy") +
  theme_minimal()

# Compare to baseline (most common pitch and zone)
baseline_accuracy <- pitch_data_features %>%
  group_by(player_name) %>%
  summarise(
    most_common_pitch = names(which.max(table(pitch_type))),
    baseline_type_accuracy = max(table(pitch_type)) / n(),
    most_common_zone = names(which.max(table(zone))),
    baseline_zone_accuracy = max(table(zone)) / n()
  )

comparison <- combined_accuracy_summary %>%
  left_join(baseline_accuracy, by = c("pitcher" = "player_name")) %>%
  mutate(
    type_improvement = type_accuracy - baseline_type_accuracy,
    zone_improvement = zone_accuracy - baseline_zone_accuracy
  )

# Type model vs baseline
ggplot(comparison, aes(x = baseline_type_accuracy, y = type_accuracy)) +
  geom_point(size = 3, alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Pitch Type Model Accuracy vs. Baseline",
       x = "Baseline Accuracy",
       y = "Model Accuracy") +
  theme_minimal()

# Zone model vs baseline
ggplot(comparison, aes(x = baseline_zone_accuracy, y = zone_accuracy)) +
  geom_point(size = 3, alpha = 0.6, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Zone Model Accuracy vs. Baseline",
       x = "Baseline Accuracy",
       y = "Model Accuracy") +
  theme_minimal()
```

## Session Info

```{r session-info}
sessionInfo()
```

