---
title: "Predicting Pitch Sequences in Baseball Using Multinomial Logistic Regression"
author: "Brian Bruxvoort"
date: "Spring 2026"
institute: "Truman State University"
format:
  beamer:
    theme: Madrid
    colortheme: beaver
    fonttheme: structurebold
    navigation: horizontal
    slide-level: 2
    toc: false
    keep-tex: true
    header-includes: |
      \usepackage{booktabs}
      \usepackage{graphicx}
      \usepackage{xcolor}
      \definecolor{truman}{RGB}{64,0,128}
      \setbeamercolor{frametitle}{bg=truman, fg=white}
      \setbeamercolor{title}{fg=truman}
      \setbeamercolor{block title}{bg=truman, fg=white}
      \setbeamertemplate{footline}[frame number]
      \setbeamertemplate{navigation symbols}{}
bibliography: references.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center")

library(tidyverse)
library(nnet)
library(caret)
library(knitr)

pitch_data1     <- read.csv("pitch_data.csv")
pitcher_models  <- readRDS("pitcher_models_basic.rds")
pitch_data_features <- readRDS("pitch_data_features_basic.rds")
model_evaluations   <- readRDS("model_evaluations_basic.rds")
model1_evals <- readRDS("model_evaluations_simple.rds")
model2_evals <- readRDS("model_evaluations_simple_stand.rds")
model3_evals <- readRDS("model_evaluations_basic.rds")
model4_type_evals <- readRDS("type_model_evaluations_separate.rds")
model4_zone_evals <- readRDS("zone_model_evaluations_separate.rds")
model5_evals <- readRDS("interaction_model_evaluations.rds")

selected_columns <- c("pitch_type", "player_name", "zone", "stand", "balls", "strikes", "at_bat_number", "pitch_number", "n_thruorder_pitcher")
pitch_data <- pitch_data1[, selected_columns]

extract_accuracy <- function(eval_list, model_name, model_type = "Combined") {
  if (is.data.frame(eval_list)) {
    return(eval_list %>% mutate(model = model_name, model_type = model_type))
  }
  data.frame(
    pitcher  = names(eval_list),
    accuracy = sapply(eval_list, function(x) {
      if (is.list(x)) {
        if ("combined_accuracy" %in% names(x)) return(x$combined_accuracy)
        if ("accuracy"          %in% names(x)) return(x$accuracy)
        return(NA)
      } else if (is.numeric(x)) return(x) else return(NA)
    }),
    stringsAsFactors = FALSE
  ) %>%
    mutate(model = model_name, model_type = model_type) %>%
    select(model, model_type, pitcher, accuracy)
}

model1_accuracy <- extract_accuracy(model1_evals, "Model 1: Simple")
model2_accuracy <- extract_accuracy(model2_evals, "Model 2: Simple + Stand")
model3_accuracy <- extract_accuracy(model3_evals, "Model 3: Basic")
model4_type_accuracy <- extract_accuracy(model4_type_evals, "Model 4: Separate", "Type Only")
model4_zone_accuracy <- extract_accuracy(model4_zone_evals, "Model 4: Separate", "Zone Only")
model5_accuracy <- extract_accuracy(model5_evals, "Model 5: Interaction")

all_accuracies <- bind_rows(
  model1_accuracy, model2_accuracy, model3_accuracy,
  model4_type_accuracy, model4_zone_accuracy, model5_accuracy)

model4_approx_combined <- all_accuracies %>%
  filter(model == "Model 4: Separate") %>%
  select(pitcher, model_type, accuracy) %>%
  pivot_wider(names_from = model_type, values_from = accuracy) %>%
  mutate(
    accuracy   = `Type Only` * `Zone Only`,
    model      = "Model 4: Separate",
    model_type = "Combined (Approximated)") %>%
  select(model, model_type, pitcher, accuracy)

all_accuracies <- bind_rows(all_accuracies, model4_approx_combined)

accuracies_filtered <- all_accuracies %>%
  filter(!(model == "Model 4: Separate" & model_type %in% c("Type Only", "Zone Only")))

pitcher_pitch_counts <- pitch_data1 %>%
  count(player_name, name = "total_pitches")

pitcher_performance <- all_accuracies %>%
  filter(model_type == "Combined") %>%
  group_by(pitcher) %>%
  summarise(
    n_models       = n(),
    mean_accuracy  = mean(accuracy, na.rm = TRUE),
    sd_accuracy    = sd(accuracy,   na.rm = TRUE),
    min_accuracy   = min(accuracy,  na.rm = TRUE),
    max_accuracy   = max(accuracy,  na.rm = TRUE),
    range_accuracy = max_accuracy - min_accuracy,
    .groups = "drop") %>%
  left_join(pitcher_pitch_counts, by = c("pitcher" = "player_name")) %>%
  arrange(desc(mean_accuracy))
```

# \color{truman}PART 1: NON-TECHNICAL {.plain}

## The Battle: Pitcher vs. Hitter

\begin{columns}
\begin{column}{0.55\textwidth}
\textbf{Every pitch is a decision:}
\begin{itemize}
  \item What type of pitch to throw?
  \item Where to throw it?
  \item What did I just throw? What's the count?
\end{itemize}

\vspace{0.4cm}
\textbf{Why does it matter?}
\begin{itemize}
  \item A 95 mph fastball reaches home plate in \textbf{\textasciitilde400 milliseconds}
  \item Knowing what's coming = precious extra reaction time
  \item The ability to anticipate a pitch can shift the competitive balance
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\begin{block}{The Core Question}
\centering
\textit{"What pitch will the pitcher throw next?"}
\end{block}
\vspace{0.3cm}
People have tried to answer this for a long time. From reading hand signals to high-tech scanning systems
\end{column}
\end{columns}

## Baseball Has a Memory

:::::: {.columns}
::: {.column width="55%"}
\textbf{MLB's Statcast system tracks everything:}
\begin{itemize}
  \item Every pitch in every game, all season long
  \item \textbf{712,528 pitches} recorded in 2025
  \item Velocity, spin, location, outcome were all captured
\end{itemize}

\vspace{0.2cm}
\textbf{Pitchers have habits:}
\begin{itemize}
  \item After a fastball, what comes next?
  \item Does the count change behavior?
  \item Does the batter's handedness matter?
\end{itemize}

\vspace{0.2cm}
\textit{Just like people have habits in daily life, pitchers develop tendencies over thousands of pitches.}
:::
::: {.column width="44%"}
```{r fig.height=3.2, fig.width=3.2}
pitch_type_summary <- pitch_data %>%
  count(pitch_type) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  arrange(desc(n))

ggplot(pitch_type_summary, aes(x = reorder(pitch_type, -n), y = n, fill = pitch_type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(title = "Pitches Thrown in 2025", x = "Pitch Type", y = "Count") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::
::::::

## The Approach: Building a Prediction Model

\begin{columns}
\begin{column}{0.6\textwidth}
\textbf{Inputs to the model:}
\begin{itemize}
  \item What were the last 1-2 pitches thrown?
  \item What is the current count (balls \& strikes)?
  \item Does the batter bat left- or right-handed?
\end{itemize}

\vspace{0.2cm}
\textbf{Output:}
\begin{itemize}
  \item A probability for each pitch type
  \item e.g., \textit{"40\% fastball, 35\% slider, 25\% changeup"}
  \item The model picks the \textbf{most likely} one as its prediction
\end{itemize}

\vspace{0.2cm}
\textbf{Key design choice:} I built a \textit{separate} model for each of the 767 pitchers in 2025 - not one generic model
\end{column}
\begin{column}{0.4\textwidth}
\begin{block}{How to think about it}
\centering
\vspace{0.2cm}
\textbf{Situation} \\
(count, last pitch, batter side)

\vspace{0.15cm}
$\Downarrow$

\vspace{0.15cm}
\textbf{Model learns pitcher's habits}

\vspace{0.15cm}
$\Downarrow$

\vspace{0.15cm}
\textbf{Predicted next pitch} \\
\textbf{probability}
\vspace{0.2cm}
\end{block}
\end{column}
\end{columns}

## What Did I Find?

:::::: {.columns}
::: {.column width="55%"}
\textbf{Prediction Accuracy:}
\begin{itemize}
  \item \textbf{62\%} correct predictions vs a 32\% random-guess baseline
  \item That's roughly \textbf{double} what you'd expect by chance
\end{itemize}

\vspace{0.3cm}
\textbf{Key finding \#1:}

Relief pitchers (closers, setup men) are typically \textbf{more predictable} than starting pitchers

\vspace{0.3cm}
\textbf{Key finding \#2:}

The \textbf{count} (balls \& strikes) is one of the biggest drivers of what pitch comes next

\vspace{0.3cm}
\textbf{Key finding \#3:}

Model complexity can backfire as more isn't always better
:::
::: {.column width="44%"}
```{r fig.height=3.5, fig.width=3.5}
overall_model_stats <- accuracies_filtered %>%
  mutate(model_combined = ifelse(
    model == "Model 4: Separate",
    paste(model, "-", model_type), model)) %>%
  filter(!model_combined %in% c(
    "Model 4: Separate - Type Only",
    "Model 4: Separate - Zone Only")) %>%
  group_by(model_combined) %>%
  summarise(mean_accuracy = mean(accuracy, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_accuracy)

ggplot(overall_model_stats,
       aes(x = reorder(model_combined, mean_accuracy), y = mean_accuracy)) +
  geom_col(fill = "purple", alpha = 0.85) +
  geom_hline(yintercept = 0.32, linetype = "dashed", color = "red", size = 0.7) +
  annotate("text", x = 0.6, y = 0.34, label = "Baseline\n(32%)", size = 2.5, color = "red") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Accuracy by Model", x = NULL, y = "Mean Accuracy") +
  theme_minimal(base_size = 8)
```
:::
::::::

## The App: Prediction in Real Time

\begin{columns}
\begin{column}{0.52\textwidth}
\textbf{The Shiny App puts the model to work:}
\begin{itemize}
  \item Select any 2025 MLB pitcher
  \item Enter the game situation (count, last pitch, batter side)
  \item Get probabilities for every pitch type
\end{itemize}

\vspace{0.2cm}
\textbf{What the app shows you:}
\begin{itemize}
  \item Predicted pitch type and confidence level
  \item Full probability breakdown for all pitch types
  \item How accurate the model is for that specific pitcher
\end{itemize}
\end{column}
\begin{column}{0.45\textwidth}
\begin{block}{App Demo}
\centering
\vspace{0.3cm}
Watch a \textbf{live video} of a pitcher and follow along pitch-by-pitch using the app.
\vspace{0.3cm}
\end{block}
\vspace{0.3cm}
\begin{block}{Practical Uses}
\begin{itemize}
  \item Pre-game scouting reports
  \item Hitter preparation
  \item Pitcher self-assessment
\end{itemize}
\end{block}
\vspace{0.2cm}
\textbf{Access the app:}\\
\footnotesize\url{https://b-bruxvoort.shinyapps.io/pitch_type_prediction_app/}
\end{column}
\end{columns}

## App Cheat Sheet: Zones & Pitch Types

\begin{columns}
\begin{column}{0.48\textwidth}
\begin{center}
\includegraphics[width=0.85\textwidth]{pitch_zones.png}

\vspace{0.1cm}
{\footnotesize \textit{Catcher/umpire's point of view}}
\end{center}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Pitch Type Abbreviations:}

\vspace{0.2cm}
{\footnotesize
\begin{tabular}{ll}
\textbf{CH} & Changeup \\
\textbf{CU} & Curveball \\
\textbf{FC} & Cutter \\
\textbf{EP} & Eephus \\
\textbf{FF} & Four-Seam Fastball \\
\textbf{FO} & Forkball \\
\textbf{KN} & Knuckleball \\
\textbf{KC} & Knuckle-curve \\
\textbf{SC} & Screwball \\
\textbf{SI} & Sinker \\
\textbf{SL} & Slider \\
\textbf{SV} & Slurve \\
\textbf{FS} & Splitter \\
\textbf{ST} & Sweeper \\
\end{tabular}
}
\end{column}
\end{columns}

## Questions

\begin{center}
\vspace{0.5cm}
{\Large\color{truman}\textbf{Questions?}}

\vspace{0.6cm}
{\large Part 1 — General Overview}

\vspace{0.8cm}
\begin{block}{}
\centering
Thank you for your time and interest.\\[0.2cm]
\textit{We will now transition into the technical presentation.}
\end{block}
\end{center}

# \color{truman}PART 2: TECHNICAL DEEP DIVE {.plain}

## Research Objectives

\textbf{Four core contributions:}

\begin{enumerate}
  \item \textbf{Systematic model comparison} — Five multinomial logistic regression specifications of increasing complexity, evaluated against one another with formal statistical tests
  \item \textbf{Pitcher-specific models} — 767 individual models rather than a single league-wide model, capturing individual pitching tendencies
  \item \textbf{Practical variable restriction} — Only variables observable in real time during a live game; no spin axis, release height, or trajectory features
  \item \textbf{Interpretability emphasis} — Logistic regression over black-box alternatives, with coefficient interpretability as a design priority
\end{enumerate}

\vspace{0.3cm}
\begin{block}{Guiding question}
Can classical statistical methods effectively model sequential pitch selection while remaining implementable in real-world game contexts?
\end{block}

## Data: Statcast 2025

:::::: {.columns}
::: {.column width="55%"}
\textbf{Source:} MLB BaseballSavant

\textbf{Scope:} 2025 regular season, all 30 teams

\textbf{Scale:} \textbf{712,528} pitch observations, 118 raw variables

\vspace{0.1cm}
\textbf{Variables retained:}
\begin{itemize}
  \item \texttt{pitch\_type} — outcome variable (14 types)
  \item \texttt{zone} — pitch location (14 zones)
  \item \texttt{stand} — batter handedness (L/R)
  \item \texttt{balls}, \texttt{strikes} — count state
  \item \texttt{at\_bat\_number}, \texttt{pitch\_number}, \texttt{player\_name}
\end{itemize}

\vspace{0.1cm}
\textbf{Why not spin rate, velocity, pfx?}

\footnotesize Multicollinearity, not real-time observable, overfitting risk
:::
::: {.column width="42%"}
```{r fig.height=3.8, fig.width=3.5}
stand_summary <- pitch_data %>%
  count(stand) %>%
  mutate(percentage = n / sum(n) * 100,
         stand_label = case_when(
           stand == "R" ~ "Right-Handed",
           stand == "L" ~ "Left-Handed",
           TRUE ~ as.character(stand))) %>% arrange(desc(n))

ggplot(stand_summary, aes(x = "", y = n, fill = stand_label)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(stand_label, "\n", n, "\n(", round(percentage, 1), "%)")), position = position_stack(vjust = 0.5), size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Left-Handed" = "#2E86AB", "Right-Handed" = "#A23B72")) +
  labs(title = "Batter Handedness\nDistribution") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5), legend.position = "none")
```
:::
::::::

## Exploratory Data Analysis: Count Distribution
```{r fig.height=2.8, fig.width=4.5}
pitch_data %>%
  count(balls, strikes) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ggplot(aes(x = factor(strikes), y = factor(balls), fill = n)) +
  geom_tile(color = "white", size = 0.8) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            color = "white", size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#3B4CC0", high = "#B40426") +
  labs(title = "Count Distribution", x = "Strikes", y = "Balls") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", panel.grid = element_blank())
```

\footnotesize \textbf{Key takeaway:} 0-0 count = 25.7\% of all pitches. Every PA starts here, making it the single most common game state.

## Exploratory Data Analysis: Pitch Sequences
```{r fig.height=3.8, fig.width=7, out.width="100%"}
pitch_sequences <- pitch_data1 %>%
  arrange(pitcher, game_date, at_bat_number, pitch_number) %>%
  group_by(pitcher, game_date, at_bat_number) %>%
  mutate(prev_pitch = lag(pitch_type)) %>%
  ungroup() %>%
  filter(!is.na(prev_pitch)) %>%
  count(prev_pitch, pitch_type) %>%
  group_by(prev_pitch) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

top_sequences <- pitch_sequences %>%
  arrange(desc(n)) %>%
  head(15) %>%
  mutate(sequence = paste(prev_pitch, "->", pitch_type))

ggplot(top_sequences, aes(x = reorder(sequence, n), y = n, fill = sequence)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(n, " (", round(percentage, 1), "%)")),
            hjust = -0.1, size = 2.8) +
  coord_flip() +
  labs(title = "Top 15 Most Common Pitch Sequences",
       x = "Pitch Sequence", y = "Count") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 10)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

\footnotesize \textbf{Key takeaway:} FF$\rightarrow$FF occurs 43.1\% of the time after a fastball as pitchers repeat their primary pitch most of the time.

## Methodological Framework

\textbf{Why Multinomial Logistic Regression?}

For an outcome $Y_i \in \{1, \ldots, J\}$ (pitch types), the softmax formulation:

$$P(Y_i = j \mid \mathbf{X}_i) = \frac{\exp(\mathbf{X}_i \boldsymbol{\beta}_j)}{\sum_{k=1}^{J} \exp(\mathbf{X}_i \boldsymbol{\beta}_k)}$$

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Advantages:}
\begin{itemize}
  \item Natural fit for unordered categorical outcomes
  \item Coefficients interpretable as log-odds ratios
  \item Computationally efficient at scale
  \item Probabilities guaranteed to sum to 1
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Why not tree-based / neural methods?}
\begin{itemize}
  \item Random forest and XGBoost tested but had no meaningful accuracy gain for computational power needed
  \item Pitch selection is approximately linear in observable game state
  \item Logistic regression is faster, interpretable, and deployable
\end{itemize}
\end{column}
\end{columns}

## The Key Addition: Lagged Variables

\textbf{Pitch selection is inherently sequential.} Pitchers "set up" batters across multiple pitches.

\begin{columns}
\begin{column}{0.5\textwidth}
\small\textbf{Lagged features created:}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parskip}{0pt}\small
  \item \texttt{lag1\_pitch\_type} — previous pitch type
  \item \texttt{lag2\_pitch\_type} — two pitches ago
  \item \texttt{lag1\_zone} — previous pitch location
  \item \texttt{lag2\_zone} — location two pitches ago
\end{itemize}
\vspace{0.05cm}
\small\textbf{At-bat boundary handling:}
\small Lag variables reset at the start of each at-bat. First pitch of a new PA receives \texttt{"None"} for all lags.
\end{column}
\begin{column}{0.47\textwidth}
\vspace{-0.4cm}
\begin{block}{Model-building code (Model 3)}
\vspace{-0.1cm}
\footnotesize
\texttt{multinom(}\\
\texttt{\ \ pitch\_type \textasciitilde}\\
\texttt{\ \ prev\_pitch\_type\_1 +}\\
\texttt{\ \ prev\_zone\_1 +}\\
\texttt{\ \ prev\_pitch\_type\_2 +}\\
\texttt{\ \ prev\_zone\_2 +}\\
\texttt{\ \ balls + strikes + stand,}\\
\texttt{\ \ maxit = 500)}
\vspace{-0.1cm}
\end{block}
\vspace{-0.1cm}
\footnotesize Fit per pitcher with $\geq 100$ pitches. MLE via \texttt{nnet::multinom()}.
\end{column}
\end{columns}

## Five Model Specifications

\begin{center}
\small
\begin{tabular}{lll}
\toprule
\textbf{Model} & \textbf{Predictors} & \textbf{Complexity} \\
\midrule
1: Simple & lag1/lag2 pitch type, pitch number & 1 \\
2: Simple $+$ Stand & Model 1 $+$ batter handedness & 2 \\
3: Basic (full) & Model 2 $+$ balls, strikes, lag1/lag2 zone & 4 \\
4: Separate & Two models — type only \& zone only & 5 \\
5: Interaction & Joint pitch type $\times$ zone outcome & 8 \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3cm}
\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Development order:}\\
Model 3 built first (kitchen sink), then Models 4/5 tested alternative outcome structures, then Models 1/2 tested whether simpler approaches were competitive.
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Model 5 rationale}
Pitch type and location decisions are correlated, but can we predict them jointly as a single outcome (e.g., "slider in zone 14")?\\
\footnotesize \textit{Issue: this creates 100$+$ outcome categories and collapses under data sparsity.}
\end{block}
\end{column}
\end{columns}

## Overall Results: Model Performance (Summary)
```{r}
overall_model_stats <- accuracies_filtered %>%
  mutate(model_combined = ifelse(
    model == "Model 4: Separate",
    paste(model, "-", model_type), model)) %>%
  filter(!model_combined %in% c(
    "Model 4: Separate - Type Only",
    "Model 4: Separate - Zone Only")) %>%
  group_by(model_combined) %>%
  summarise(
    Mean  = round(mean(accuracy, na.rm = TRUE), 4),
    SD    = round(sd(accuracy,   na.rm = TRUE), 4),
    .groups = "drop") %>%
  arrange(desc(Mean)) %>%
  rename(Model = model_combined)

knitr::kable(overall_model_stats, format = "latex", booktabs = TRUE)
```

\vspace{0.2cm}
\footnotesize \textbf{Key:} Model 3 = 62.36\%; Model 1 = 47.20\%; Model 5 = 23.10\% (below baseline).

## Overall Results: Model Performance (Distribution)
```{r fig.height=3.5, fig.width=7, out.width="100%"}
ggplot(accuracies_filtered,
       aes(x = reorder(paste(model, model_type), accuracy, FUN = median),
           y = accuracy, fill = model)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Accuracy Distribution by Model", x = NULL, y = "Accuracy") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

\footnotesize Model 3's wide distribution signals strong pitcher-to-pitcher variation meaning it excels for predictable pitchers and struggles for unpredictable ones.

## Statistical Significance Testing

```{r, include=FALSE}
accuracy_comparison <- accuracies_filtered %>%
  filter(model_type %in% c("Combined", "Combined (Approximated)")) %>%
  select(model, pitcher, accuracy)

common_pitchers <- accuracy_comparison %>%
  group_by(pitcher) %>%
  summarise(n_models = n_distinct(model), .groups = "drop") %>%
  filter(n_models >= 4) %>%
  pull(pitcher)

accuracy_comparison <- accuracy_comparison %>%
  filter(pitcher %in% common_pitchers)

anova_result <- aov(accuracy ~ model, data = accuracy_comparison)
anova_sum <- summary(anova_result)
tukey_result <- TukeyHSD(anova_result)
```

\textbf{ANOVA:} $F(4, 3659) = 483.2$, $p < 2 \times 10^{-16}$ which shows overwhelming evidence of differences

\vspace{0.3cm}
\textbf{Tukey HSD — Notable Pairwise Comparisons:}

\small
\begin{center}
\begin{tabular}{lrrl}
\toprule
\textbf{Comparison} & \textbf{Mean Diff} & \textbf{Cohen's d} & \textbf{Effect} \\
\midrule
M1 vs M2 (add handedness) & $+$2.54 pp & $-0.232$ & Small \\
M1 vs M3 (add count$+$zone) & $+$15.16 pp & $-1.011$ & Large \\
M2 vs M3 (isolate count$+$zone) & $+$12.62 pp & $-0.853$ & Large \\
M3 vs M4 (joint prediction cost) & $+$28.04 pp & $+1.122$ & Large \\
M3 vs M5 (interaction failure) & $+$39.26 pp & $+3.413$ & Enormous \\
M4 vs M5 (separate vs joint) & $+$11.22 pp & $+0.428$ & Medium \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2cm}
\footnotesize All 10 pairwise comparisons significant at $\alpha = 0.01$. Count and location contribute the largest marginal gains; Model 5 failure is structurally severe.

## Why Model 5 Failed: Data Sparsity

\begin{columns}
\begin{column}{0.52\textwidth}
\textbf{The problem:}
\begin{itemize}
  \item $7$–$10$ pitch types $\times$ $14$ zones $= 98$–$140$ outcome categories per pitcher
  \item A pitcher with 2,500 pitches may produce 80–100 unique type-zone combos
  \item Many combos appear only 5–10 times which is insufficient for stable MLE
\end{itemize}

\vspace{0.15cm}
\footnotesize \textbf{Lesson:} More information can hurt when complexity grows faster than data availability. The theoretical motivation was sound; the practical execution was not.
\end{column}
\begin{column}{0.45\textwidth}
\begin{block}{Why Model 4 fares better}
\small Model 4 separates the two predictions. Each sub-model has a manageable number of outcome categories (14 pitch types \textit{or} 14 zones). Combined accuracy is approximated as:
$$\hat{A}_{\text{combined}} = \hat{A}_{\text{type}} \times \hat{A}_{\text{zone}}$$
The multiplicative penalty still produces low combined accuracy (34.32\%), but the individual models are statistically tractable.
\end{block}
\end{column}
\end{columns}

## Pitcher-Specific Analysis: The Relief vs. Starter Divide

:::::: {.columns}
::: {.column width="48%"}
\textbf{Most predictable (min. 500 pitches):}
```{r}
pitcher_performance %>%
  filter(total_pitches >= 500) %>%
  head(8) %>%
  select(pitcher, mean_accuracy, total_pitches) %>%
  mutate(mean_accuracy = round(mean_accuracy, 3)) %>%
  rename(Pitcher = pitcher, Acc = mean_accuracy, N = total_pitches) %>%
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "")
```
:::
::: {.column width="48%"}
\textbf{Least predictable:}
```{r}
pitcher_performance %>%
  filter(total_pitches >= 500) %>%
  tail(8) %>%
  select(pitcher, mean_accuracy, total_pitches) %>%
  mutate(mean_accuracy = round(mean_accuracy, 3)) %>%
  arrange(mean_accuracy) %>%
  rename(Pitcher = pitcher, Acc = mean_accuracy, N = total_pitches) %>%
  knitr::kable(format = "latex", booktabs = TRUE, linesep = "")
```
:::
::::::

\vspace{0.1cm}
\footnotesize \textbf{Pattern:} Most predictable = relief pitchers (Kahnle, Jansen, Hill). Least predictable = experienced starters (Lugo, Darvish, Alcantara). \textbf{Elite starter unpredictability is a skill, not a modeling artifact.}

## Pitcher Predictability Distribution

:::::: {.columns}
::: {.column width="55%"}
```{r fig.height=3.5, fig.width=4}
ggplot(pitcher_performance, aes(x = mean_accuracy)) +
  geom_histogram(binwidth = 0.02, fill = "purple", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mean_accuracy)), color = "red", linetype = "dashed", size = 0.8) +
  annotate("text", x = mean(pitcher_performance$mean_accuracy) + 0.005, y = Inf, vjust = 2, label = paste0("Mean = ", round(mean(pitcher_performance$mean_accuracy), 3)), color = "red", size = 2.8) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Distribution of Mean Accuracy Across Pitchers", x = "Mean Accuracy", y = "# Pitchers") +
  theme_minimal(base_size = 9)
```
:::
::: {.column width="42%"}
\textbf{Key statistics:}
\begin{itemize}
  \item Overall mean: \textbf{48.09\%}
  \item Median: 45.60\%
  \item SD: 13.53\%
  \item Range: \textasciitilde25\% to 70\%$+$
\end{itemize}

\vspace{0.3cm}
\textbf{Accuracy--SD correlation:}

$r = 0.566$ ($p < 0.001$)

\vspace{0.2cm}
More predictable pitchers show \textit{greater model-to-model variance}. When real patterns exist, different models capture different aspects of those patterns. When no patterns exist, all models fail similarly.
:::
::::::

## Model Efficiency: Complexity vs. Accuracy

:::::: {.columns}
::: {.column width="55%"}
```{r fig.height=3.5, fig.width=4.5}
model_complexity <- data.frame(
  Model = c("Model 1: Simple",
            "Model 2: Simple + Stand",
            "Model 3: Basic",
            "Model 4: Separate - Combined (Approximated)",
            "Model 5: Interaction"),
  Num_Features = c(3, 4, 7, 7, 10),
  Complexity_Score = c(1, 2, 4, 5, 8))

overall_model_stats_eff <- accuracies_filtered %>%
  mutate(model_combined = ifelse(
    model == "Model 4: Separate",
    paste(model, "-", model_type), model)) %>%
  filter(!model_combined %in% c(
    "Model 4: Separate - Type Only",
    "Model 4: Separate - Zone Only")) %>%
  group_by(model_combined) %>%
  summarise(mean_accuracy = mean(accuracy, na.rm = TRUE), .groups = "drop")

complexity_performance <- model_complexity %>%
  left_join(overall_model_stats_eff, by = c("Model" = "model_combined")) %>%
  mutate(efficiency = mean_accuracy / Complexity_Score)

ggplot(complexity_performance,
       aes(x = reorder(Model, efficiency), y = efficiency, fill = Model)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sprintf("%.4f", efficiency)),
            hjust = -0.1, size = 3, fontface = "bold") +
  coord_flip() +
  labs(title = "Model Efficiency: Accuracy per Complexity Unit",
       subtitle = "Higher values = better performance relative to complexity",
       x = NULL, y = "Accuracy / Complexity Score") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 9),
        plot.subtitle = element_text(size = 7),
        legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```
:::
::: {.column width="42%"}
\small\textbf{Recommendations by use case:}

\vspace{0.1cm}
\begin{block}{Max Accuracy}
\small\textbf{Model 3} — 62.36\%, large effect sizes justify complexity
\end{block}

\vspace{0.05cm}
\begin{block}{Best Efficiency}
\small\textbf{Model 1} — 47.20\%, minimal computation, easy to deploy
\end{block}

\vspace{0.05cm}
\begin{block}{Joint Type + Location}
\small\textbf{Model 4 (Separate)} — only viable approach; 34.32\%
\end{block}

\vspace{0.05cm}
\begin{alertblock}{Avoid}
\small\textbf{Model 5} — 23.10\%, worst accuracy at highest complexity
\end{alertblock}
:::
::::::

## Limitations

\begin{columns}
\begin{column}{0.48\textwidth}
\small\textbf{Data \& temporal:}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parskip}{0pt}\small
  \item Single season (2025) only meaning there was no temporal stability assessment
  \item Static models don't capture within-season adjustments (fatigue, injury, opponent adaptation)
\end{itemize}
\vspace{0.05cm}
\small\textbf{Methodological:}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parskip}{0pt}\small
  \item No formal cross-validation, but relative comparisons remain valid even if absolute accuracy is slightly optimistic
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\small\textbf{Contextual:}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parskip}{0pt}\small
  \item Broad game context excluded: score differential, base-out state, inning, bullpen availability
  \item Catcher influence ignored. Catchers significantly shape pitch selection, especially for younger pitchers
\end{itemize}
\vspace{0.05cm}
\small\textbf{Scope:}
\begin{itemize}
  \setlength{\itemsep}{1pt}\setlength{\parskip}{0pt}\small
  \item No field experiment meaning we cannot establish that predictions improve offensive performance
  \item Results are correlational, not causal
\end{itemize}
\end{column}
\end{columns}

## Future Research Directions

\begin{columns}
\begin{column}{0.48\textwidth}
\textbf{Modeling improvements:}
\begin{itemize}
  \item \textbf{Multi-season analysis} — do pitcher patterns persist across years?
  \item \textbf{Hierarchical Bayesian models} — let pitchers "borrow strength" from similar pitchers
  \item \textbf{Recurrent neural networks} — longer sequence memory beyond a two-pitch lookback
  \item \textbf{Hidden Markov Models} — uncover latent pitcher "modes" (aggressive vs. conservative)
\end{itemize}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Substantive research questions:}
\begin{itemize}
  \item Does predictability correlate with ERA / WAR?
  \item Do patient hitters exploit predictable pitchers more than aggressive hitters?
  \item Can predictability changes serve as an early injury indicator?
  \item Can we add base-out state, score differential, inning, or more?
\end{itemize}
\vspace{-0.1cm}
\begin{block}{Possible Future Goal}
\small Controlled field experiment where we deploy predictions to minor league hitters and measure offensive outcome changes
\end{block}
\end{column}
\end{columns}

## Conclusion

\begin{columns}
\begin{column}{0.48\textwidth}
\begin{enumerate}
  \item \textbf{Game state drives selection.} Count is the dominant predictor as adding it yielded a 12.62 pp gain ($d = -0.853$)
  \item \textbf{Role predicts predictability.} Relief pitchers are more predictable; elite starter unpredictability is a cultivated skill
\end{enumerate}
\end{column}
\begin{column}{0.48\textwidth}
\begin{enumerate}\setcounter{enumi}{2}
  \item \textbf{Complexity must match data.} Model 5's failure ($d = 3.413$) shows what happens when categories outgrow the data
  \item \textbf{Classical methods are competitive.} MLR at 62.36\% matches neural network benchmarks while staying interpretable
\end{enumerate}
\end{column}
\end{columns}

\vspace{0.2cm}
\begin{block}{Broader applicability}
\small Categorical lagged variables in multinomial logistic regression can be applied beyond baseball where behavior unfolds in structured categorical sequences like healthcare, digital behavior, and other sequential decision systems.
\end{block}

\vspace{0.05cm}
\footnotesize Thesis advisor: Dr. Scott Alberts\\
Department of Computer and Data Sciences\\
Truman State University, 2026

## Questions

\begin{center}
\vspace{0.5cm}
{\Large\color{truman}\textbf{Questions?}}

\vspace{0.6cm}
{\large Part 2 — Technical Deep Dive}

\vspace{0.6cm}
\begin{block}{}
\centering
Thank you to Dr.\ Alberts and my committee\\
for their guidance throughout this research.\\[0.3cm]
\textit{Full thesis, code, and app available upon request.}
\end{block}

\vspace{0.4cm}
\footnotesize
\textbf{App:} \url{https://b-bruxvoort.shinyapps.io/pitch_type_prediction_app/}
\end{center}
